\PaperTitle

### 2020-05-17 paper/filter_bubbles.tex
finish revised except appendix

\documentclass[format=acmsmall, review=false]{acmart}
\settopmatter{printacmref=false, printccs=false}
\usepackage{booktabs} % For formal tables
\usepackage[ruled]{algorithm2e} % For algorithms

\usepackage{subcaption}
\newcommand{\xhdr}[1]{\vspace{1mm} \noindent{\bf #1}}
\renewcommand{\footnotetextcopyrightpermission}[1]{} % remove copyright block
\renewcommand{\footnotetextauthorsaddresses}[1]{} % remove author address 
\renewcommand{\algorithmcfname}{ALGORITHM}
\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}
\SetAlCapHSkip{0pt}
\IncMargin{-\parindent}

@@

We denote by $x_{i,n}$ user $i$'s realized value from consuming item $n$. In particular, we consider that the realized value derived from a given item can be decomposed in the following manner: $x_{i,n}= v_{i,n} + \beta v_n$, where $v_{i,n}$ denotes an idiosyncratic component -- i.e. user $i$'s idiosyncratic taste for product $n$ --  and $v_{n}$, a common-value component. One can interpret $v_n$ as a measure of how much item $n$ is valued in society in general and, in a sense, $v_{i,n}$ denotes how $i$ diverges from this overall ranking. The scalar $\beta \in \mathbb{R}_{t}$ denotes the degree to which valuations are idiosyncratic to each user or common across users. If $\beta=0$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta$ is large, every individual has similar preferences.
\par
Stacking values in vector-form, we get the vector of values associated with each item 
$${\left(x_{i,n}\right)}_{n \in \mathcal{J}}=:X_i =V_i+ \beta V $$
where $V_i ={\left(v_{i,n}\right)}_{n \in \mathcal{J}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{J}}$.
\par
\xhdr{User Decision-Making.}
We assume the user makes $T$ choices and therefore can only consume up to $T$ items, where $T$ is a small fraction of $N$. This captures the idea that users are faced with an immense choice set but that ultimately they end up experiencing (and learning) about just a small fraction of it. Since this is a sequential decision-making problem under uncertainty, in principle users face an exploration-exploitation trade-off. However, for tractability, we impose that users are myopic and every period consume the product that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
\par
\xhdr{User Beliefs.} We assume that all the true realized values are drawn at $t = 0$. However, users do not know the realized values before consuming an item, but rather have beliefs over the realized utilities.
Formally, user $i$ starts with some beliefs about $X_i$, namely that the idiosyncratic and common-value parts of the valuations are independent -- $V_i \perp \!\!\! \perp V$ -- and that each is multivariate normal 
\begin{enumerate}
\item $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ 
\item $V \sim \mathcal N(\overline V, \Sigma)$ with $\overline V =0$
\end{enumerate}
We impose the normality assumption for two reasons. The first is that this allows for simple and tractable belief updating. The second is that it allows us to incorporate an easily interpretable correlation structure between the items. The precise formulation of $\Sigma$ and $\Sigma_i$ that we consider is defined below when we discuss user learning.
\par
Keeping with the assumption that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{i,n}\right)_{n \in \mathcal{J}}$ are drawn independently from a jointly normal distribution, where $\overline v_{i,n} \sim \mathcal N (0, \overline \sigma^2)$ are independent and identically distributed. These $\overline v_{i,n}$ denote the prior belief that individual $i$ holds about her valuation over item $n$. As people are exposed to different backgrounds, their beliefs about how much they value a given item also varies and $\overline v_{i,n}$ denotes this idiosyncrasy at the level of prior beliefs.
\par

@@

\xhdr{An Illustrative Example} We illustrate the main intuitions of our model with a simple example. Suppose that there are four items: 0, 1, 2, 3. The items are in different places of the product space, where 0 is close to 1 and 3 but more distant from 2, as shown in Figure \ref{fig:illustrative_example}. The initial beliefs are given as follows:

@@

We will consider three cases. The case of primary interest is \textit{recommendation} where the recommender observes values accrued and knows $V$ but does not know $V_i$.\footnote{We do not consider the acquisition of information for the recommender to know $V$ and suppose that she has sufficient data to learn $V$ with arbitrary precision at $t = 0$.} However, the recommender does know the users' beliefs $\bar V_i$. Thus, at any given period, the recommender provides a personalized recommendation that combines the knowledge of the common value component $V$ with the user beliefs $\bar V_i$. Knowing the user's beliefs about her valuation of each item become crucial in this case: just providing the user with information about $V$ may change the user's original ranking, but, without considering the user's beliefs, she will not necessarily follow the recommendation.\footnote{The notion of recommendation that we consider is idealized where the recommendation does the Bayesian updating for users. As users face a very difficult and costly Bayesian updating instance, a useful recommendation system provides recommendations taking into account this aspect and combines the history of past consumption and value realizations with the user's beliefs. 
An alternative that is equivalent would be the recommender simply providing $V$ without knowing $V_i$. In this case, the recommendation will be that the user $i$ chooses $r_{i,t} \in \arg \max_{n\in \mathcal J\backslash C^T_i} \mathbb E[u(x_n) \mid C_i^T]$
but we assume the recommender provides full information about $V$. For instance, the recommender could display the whole distribution of utilities reported by other users or even its average, which is a good proxy for the common value component. It is possible for the recommender to only recommend the best item given such information, that is, the item with highest common value component. However, in that case, recommending only a single item generates costly Bayesian updating to the user and provides little guidance as to what she should indeed pick as the common value might be of little importance when compared to the idiosyncratic component. Therefore, we assume that the recommender reports the whole $V$, which results in higher expected welfare in choices and is not costly to implement.
The results would be equivalent, but this would shift the burden of Bayesian updating to users.}
\par

@@

We simulate over risk-aversion parameters $\gamma \in \{ 0, 0.3, 0.6, 1, 5 \}$, \ standard-deviation $\sigma \in \{ 0.25, 0.5, 1.0, 2.0, 4.0 \}$, correlation parameters $\rho\in \{ 0, 0.1, 0.3, 0.5, 0.7, 0.9 \} $ and degree of idiosyncrasy of individual valuations $\beta \in \{ 0, 0.4, 0.8, 1, 2, 5\}$. The range of considered parameter values cover the relevant portions of the parameter space in order to provide full insight into the behavior of the model. 
\par
When we consider results varying a single parameter, we group the results over the other parameters and provide reports varying only the parameter of interest. We report results for a relatively small consumption history $T=20$ with a product space size $N=200$.
\par

@@

First, the right panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho = 0$, there is no difference in consumption distance between the three regimes. This is due to the fact that when $\rho = 0$, there is no reason that items that are close in the product space should have similar values and so the optimal consumption path does not depend on the similarity of the items. However this also means that users do not learn anything about neighboring products and so there is limited path-dependence in consumption. Not only is there no difference in the levels between the three regimes, but they all have the same, flat, average consumption distance path. This underscores the fact that such concerns about filter bubbles \textit{only} arise in a world where there is correlation between the realized utility of items. If there were no correlation between the realized utilities then there would be no reason for users to consume similar items and thus no narrowing effect, regardless of the information on the true utility of the items that users had.
\par
The top panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho > 0$, both recommendation and no recommendation lead to increasingly local consumption compared to the oracle benchmark case. Further, the average consumption path between periods is \textit{decreasing} for the no recommendation case whereas it is \textit{increasing} for the oracle case. The recommendation regime decreases the degree of local consumption, but not as much as the oracle benchmark. Due to the correlation of values, the oracle consumption path exploits this and leads to the consumption of more similar products than in the case when $\rho = 0$. However, since these spillovers also impact user learning in the no recommendation case, users \textit{over-exploit} this and increasingly consume products similar to good products that they have consumed before. This is further illustrated by Figure \ref{fig:local_consumption_across_rho} which shows how the consumption paths in the oracle and no recommendation regimes vary as $\rho$ increases.
\par

Finally, this effect is further amplified as the level of risk aversion, $\gamma$, increases. Figure \ref{fig:no_rec_risk_aversion} shows how drastically the degree of local consumption increases as $\gamma$ increases. This is due to the fact that the spillovers not only affect the mean expected belief about quality but also the degree of uncertainty. Local consumption therefore leads to users having less uncertainty about certain areas of the product space and risk aversion may lead them to increasingly consume nearby products.

@@

\begin{finding}\label{finding_diversity}
The impact of recommendation on product diversity:
\begin{enumerate}
\item When $\rho = 0$, product diversity is the same across all three recommendation regimes;
\item When $\rho > 0$, product diversity decreases across all recommendation regimes but decreases the most in the no recommendation regime. This effect is amplified as $\rho$ increases as well as when users become more risk-averse.
\end{enumerate}
\end{finding}
\par 
Finding~\ref{finding_diversity} summarizes the main results on product diversity. As before, when there is no correlation between valuations, product diversity is the same across different recommendation regimes. The over-exploitation of information spillovers when $\rho > 0$ leads to product diversity being lowest in the no recommendation regime. As a result, this effect gets amplified as $\rho$ increases, which leads to the gap in diversity between the regimes to increase as $\rho$ increases. Figure \ref{fig:diversity_welfare_correlation} shows how diversity varies as $\rho$ increases across the three regimes that we consider. There is a similar increasing diversity gap as $\gamma$, or the level of risk-aversion, increases as can be seen in Figure \ref{fig:risk_aversion_diversity}. The mechanisms behind these effects have direct parallels to those discussed in the previous section since low average sequential consumption distance directly leads to low diversity.
\par 
We now study how recommendation impacts user welfare in our model. In our model users make consumption decisions that maximize their current period ex-ante utility that depends on their beliefs in that period. Thus, from an ex-ante perspective, they make optimal decisions, but our primary interest is in understanding how the ex-post, or realized, utility varies across regimes and parameter values. We define user's \textit{ex-post} welfare as the average of the realized values, controlling for the effect of $T$:
$$W_i:= \frac{1}{T}\sum_{n \in C_i^T} x_{i,n}$$

\begin{finding}\label{finding_welfare_gap}
The impact of recommendation on consumer welfare is as follows:
\begin{enumerate}
\item Under oracle recommendation, welfare is constant as $\rho$ increases.
\item Under no recommendation, welfare is increasing as $\rho$ increases.
\item Recommendation introduces welfare gains relative to no recommendation, but these gains are decreasing as $\rho$ increases. 
\end{enumerate}
\end{finding}
\par 
Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which are drawn from Figure~\ref{fig:diversity_welfare_correlation}. The most interesting observation is that the value of recommendation decreases as $\rho$ decreases. The intuition for this is clear as recommendation provides users with information that allows them to better guide their decisions and increase welfare. However, as $\rho$ increases users get increasingly more information simply from consuming items since the realized utility is now more informative about the utility of nearby items. Thus, since consumption decisions themselves reveal information, the information that recommendation reveals provides less value.
\par 
Figure~\ref{fig:diversity_welfare_common_value} shows how diversity and welfare vary across recommendation regimes as the the strength of the common-value component, $\beta$, varies. As expected, when $\beta = 0$, these values are the same. However, as $\beta$ grows the idiosyncratic component of user valuations becomes increasingly less important and the recommendation regime begins to converge to the oracle regime. Indeed, we see that both product diversity and welfare increase under recommendation, whereas under no recommendation we see that there is a decrease in diversity with little changes to consumer welfare. This is intuitive as $\beta$ controls the informativeness of recommendation about user preferences.
\par 
One striking observation is that the decrease in diversity does not appear to be associated with a decline in welfare. Indeed, it appears that the opposite is the case - that low diversity is associated with higher welfare and vice versa. Thus, we next explore the relationship between welfare and diversity.
\begin{figure}
\caption{Relationship between User Welfare and Diversity, varying $\rho$}\label{fig:diversity_welfare_correlation}
\begin{subfigure}{.45\linewidth}
  \includegraphics[width=.8\linewidth]{figures/rho_diversity_N_200_T_20.pdf}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
  \includegraphics[width=.8\linewidth]{figures/rho_welfare_N_200_T_20.pdf}
\end{subfigure}
\caption*{\scriptsize Notes: The figure on the left displays the relationship between $rho$, the correlation between valuations of items, and overall consumed product diversity. The figure on the right displays the relationship between $\rho$ and overall welfare.}
\end{figure}

@@

Finding~\ref{finding_diversity_welfare_corr} summarizes our findings on the relationship between diversity and welfare. Figure \ref{fig:diversity_welfare_ra} shows how diversity and welfare correlate for the no recommendation case as we vary the degree of risk aversion. When there is no risk-aversion then there is a negative correlation between welfare and diversity. This is since, with no risk-aversion, in a given period a user will select the good that she currently believes has the highest expected value. High product diversity in this case can arise from a user who consumes an item she disliked and updates her beliefs about nearby items negatively. As a result, in the following period she will pick an item far away in the product space from the item that was previously consumed. If instead the user valued highly the item that she had consumed, then she is more likely to pick a nearby item. The information spillovers therefore lead to high product diversity being negatively correlated with welfare.
\par
This only happens since $\gamma = 0$ leads to users only caring about the expected value of the product. However, as we saw in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, increasing $\gamma$ can lead to lower diversity and increasingly local consumption due to the fact that the degree of uncertainty now impacts users' choices. This weakens the negative relationship between diversity and welfare since both negative and positive experiences with a good reduce uncertainty about surrounding items. This leads to the inverted-U shape found in Figure \ref{fig:diversity_welfare_ra} when $\gamma$ is relatively large (e.g $\gamma = 5$) though diversity and welfare are virtually uncorrelated in the data. In the recommendation and oracle regimes, under risk neutrality ($\gamma=0$), welfare and diversity and uncorrelated, while under risk aversion ($\gamma=5$), it is possible to observe in Figure \ref{fig:diversity_welfare_ra_partial} an actual positive relation between diversity and welfare as recommendations are able to reduce uncertainty and facilitate exploration of the product space.
\begin{figure}[t]
\caption{Diversity vs. Welfare, \textbf{No Recommendation}}
\begin{subfigure}{.45\textwidth}
\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn.pdf}
\end{subfigure}
\begin{subfigure}{.45\textwidth}
\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra.pdf}
\end{subfigure}\\
\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under the no recommendation regime. The figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.}\label{fig:diversity_welfare_ra}
\end{figure}

\begin{figure}[t]
\caption{Diversity vs. Welfare, \textbf{Recommendation}}
\begin{subfigure}{.45\textwidth}
\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_partial.pdf}
\end{subfigure}
\begin{subfigure}{.45\textwidth}
\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_partial.pdf}
\end{subfigure}\\
\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under the recommendation regime. The figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.}\label{fig:diversity_welfare_ra_partial}
\end{figure}

\subsection{User Homogenization}
In this section, we focus on comparisons across users and investigate how the consumed set of items across users varies across different recommendation regimes and parameter values. In particular we look at the degree of \textit{homogenization} between users. Similar to other papers that study the degree of homogenization in recommender systems (e.g. \cite{chaney2018algorithmic}) we measure homogeneity via the Jaccard index between the consumption sets of users:
\begin{align*}
H:=\frac{1}{|I|(|I|-1)}\sum_{i,j \in I: i \ne j}d_J(C_i^T,C_j^T)
\end{align*}
where $d_J$ denotes the Jaccard index and $H \in [0,1]$.
\begin{finding}\label{finding_homogeneity}
The impact of recommendation on homogeneity is as follows:
\begin{enumerate}
\item Highest under recommendation and lowest under no recommendation;
\item Increasing in $\beta$, or the weight of the common-value component;
\item Decreasing in $\rho$ for partial recommendation, but weakly increasing in $\rho$ for no recommendation.
\end{enumerate}
\end{finding}
Finding~\ref{finding_homogeneity} summarizes our findings on the impact of recommendation on user homogeneity. First, we study how the degree of homogenization varies as we increase $\beta$, the weight of the common value component. As $\beta$ increases we expect that users should become increasingly homogeneous as the realized utilities of the items are now increasingly similar. Figure \ref{fig:beta_homo} confirms that as the weight of the common-value component, $\beta$, increases users consume increasingly similar sets of items. The homogenization effect is strongest under the recommendation regime since the revelation of the common-value component induces users to consume products in similar areas of the product space. As $\beta$ increases, some amount of homogenization is optimal as can be seen from the oracle case. However, since users in the no recommendation regime do not know the common-value component they engage in local consumption in different areas of the product space which leads to less than optimal homogeneity.
\par
We next study how the degree of homogeneity varies as $\rho$ increases. Figure \ref{fig:cor_homo} shows how homogeneity varies as $\rho$ increases across the different recommendation regimes. The most striking finding is that homogeneity decreases as $\rho$ increases in the recommendation regime. As was highlighted in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, the degree of local consumption increases with $\rho$. Even though the revelation of the common-value component induces them to search in similar parts of the product space, their idiosyncratic components induce them to consume products in a more localized area of the product space as $\rho$ increases which leads to a decline in homogeneity.

@@

In this section we discuss how the insights from our model of user decision-making can inform the evaluation and design of recommender systems. The classic approach to recommender system evaluation is predicting user ratings for items and comparing how accurate this prediction is to recorded ratings data, either explicitly given by users or inferred from behavioral data. The recommender system should then recommend to users the items with the highest predicted rating \cite{adomavicius2005toward}.

>>>>

\documentclass[format=acmsmall, review=false]{acmart}
\settopmatter{printacmref=false, printccs=false}
\usepackage{booktabs} % For formal tables
\usepackage[ruled]{algorithm2e} % For algorithms
\usepackage{enumitem}
\usepackage{subcaption}
\newcommand{\xhdr}[1]{\vspace{1mm} \noindent{\bf #1}}
\renewcommand{\footnotetextcopyrightpermission}[1]{} % remove copyright block
\renewcommand{\footnotetextauthorsaddresses}[1]{} % remove author address 
\renewcommand{\algorithmcfname}{ALGORITHM}
\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}
\SetAlCapHSkip{0pt}
\IncMargin{-\parindent}

@@

We denote by $x_{i,n}$ user $i$'s realized value from consuming item $n$. In particular, we consider that the realized value derived from a given item can be decomposed in the following manner: $x_{i,n}= v_{i,n} + \beta v_n$, where $v_{i,n}$ denotes an idiosyncratic component -- i.e. user $i$'s idiosyncratic taste for product $n$ --  and $v_{n}$, a common-value component. One can interpret $v_n$ as a measure of how much item $n$ is valued in society in general and, in a sense, $v_{i,n}$ denotes how $i$ diverges from this overall ranking. The scalar $\beta \in \mathbb{R}_{t}$ denotes the degree to which valuations are idiosyncratic to each user or common across users. If $\beta=0$, it is impossible to generate meaningful predictions of any one's individual preferences based on others, while if $\beta$ is large, every individual has similar preferences.
\par
Stacking values in vector-form, we get the vector of values associated with each item 
$${\left(x_{i,n}\right)}_{n \in \mathcal{J}}=:X_i =V_i+ \beta V $$
where $V_i ={\left(v_{i,n}\right)}_{n \in \mathcal{J}}$ and $V={\left(v_{n}\right)}_{n \in \mathcal{J}}$.
\par
\xhdr{User Decision-Making.}
We assume the user makes $T$ choices and therefore can only consume up to $T$ items, where $T$ is a small fraction of $N$. This captures the idea that users are faced with an immense choice set but that ultimately they end up experiencing (and learning) about just a small fraction of it. Since this is a sequential decision-making problem under uncertainty, in principle users face an exploration-exploitation trade-off. However, for tractability, we impose that users are myopic and every period consume the product that they have not yet tried ($n_i^t$) that gives them the highest expected utility given the information from past consumption offers ($C_i^{t-1}=(n_i^1,...,n_i^{t-1})$) and their initial beliefs.
\par
\xhdr{User Beliefs.} We assume that all the true realized values are drawn at $t = 0$. However, users do not know the realized values before consuming an item, but rather have beliefs over the realized utilities.
Formally, user $i$ starts with some beliefs about $X_i$, namely that the idiosyncratic and common-value parts of the valuations are independent -- $V_i \perp \!\!\! \perp V$ -- and that each is multivariate normal 
\begin{enumerate}[topsep=0pt]
\item $V_i \sim \mathcal N (\overline V_i, \Sigma_i)$ 
\item $V \sim \mathcal N(\overline V, \Sigma)$ with $\overline V =0$
\end{enumerate}
We impose the normality assumption for two reasons. The first is that this allows for simple and tractable belief updating. The second is that it allows us to incorporate an easily interpretable correlation structure between the items. The precise formulation of $\Sigma$ and $\Sigma_i$ that we consider is defined below when we discuss user learning.
\par
Keeping with the assumption that $V_i$ represents idiosyncratic deviations from $V$, we assume that, on the population level, prior beliefs $\overline V_i=\left(\overline v_{i,n}\right)_{n \in \mathcal{J}}$ are drawn independently from a jointly normal distribution, where $\overline v_{i,n} \sim \mathcal N (0, \overline \sigma^2)$ are independent and identically distributed. These $\overline v_{i,n}$ denote the prior belief that individual $i$ holds about her valuation over item $n$. As people are exposed to different backgrounds, their beliefs about how much they value a given item also varies and $\overline v_{i,n}$ denotes this idiosyncrasy at the level of prior beliefs.
\par

@@

\xhdr{An Illustrative Example} We illustrate the main intuitions of our model with a simple example. Suppose that there are four items: 0, 1, 2, 3. The items are in different places of the product space, where 0 is close to 1 and 3 but more distant from 2. The initial beliefs are given as follows:

@@

We will consider three cases. The case of primary interest is \textit{recommendation} where the recommender observes values accrued and knows $V$ but does not know $V_i$.\footnote{We do not consider the acquisition of information for the recommender to know $V$ and suppose that she has sufficient data to learn $V$ with arbitrary precision at $t = 0$.} However, the recommender does know the users' beliefs $\bar V_i$. Thus, at any given period, the recommender provides a personalized recommendation that combines the knowledge of the common value component $V$ with the user beliefs $\bar V_i$. Knowing the user's beliefs about her valuation of each item become crucial in this case: just providing the user with information about $V$ may change the user's original ranking, but, without considering the user's beliefs, she will not necessarily follow the recommendation.
\par

@@

We simulate over risk-aversion parameters $\gamma \in \{ 0, 0.3, 0.6, 1, 5 \}$, \ standard-deviation $\sigma \in \{ 0.25, 0.5, 1.0, 2.0, 4.0 \}$, correlation parameters $\rho\in \{ 0, 0.1, 0.3, 0.5, 0.7, 0.9 \} $ and degree of idiosyncrasy of individual valuations $\beta \in \{ 0, 0.4, 0.8, 1, 2, 5\}$. The range of considered parameter values cover the relevant portions of the parameter space in order to provide full insight into the behavior of the model. 
\par
When we consider results varying a single parameter, we group the results over the other parameters and provide reports varying only the parameter of interest. We report results for a relatively small consumption history $T=20$ with a product space size $N=200$.\footnote{In the supplementary material we include the same exercises presented here for $N = 100$ and $N = 500$ and show that the results are qualitatively similar to those reported here.}
\par

@@

First, the left panel of Figure \ref{fig:correlation_consumption_path} shows that, when $\rho = 0$, there is no difference in consumption distance between the three regimes. This is due to the fact that when $\rho = 0$, there is no reason that items that are close in the product space should have similar values and so the optimal consumption path does not depend on the similarity of the items. However this also means that users do not learn anything about neighboring products and so there is limited path-dependence in consumption. Not only is there no difference in the levels between the three regimes, but they all have the same, flat, average consumption distance path. This underscores the fact that such concerns about filter bubbles \textit{only} arise in a world where there is correlation between the realized utility of items. If there were no correlation between the realized utilities then there would be no reason for users to consume similar items and thus no narrowing effect, regardless of the information on the true utility of the items that users had.
\par
The right panel of Figure~\ref{fig:correlation_consumption_path} shows that, when $\rho > 0$, both recommendation and no recommendation lead to increasingly local consumption compared to the oracle benchmark case. Further, the average consumption path between periods is \textit{decreasing} for the no recommendation case whereas it is \textit{increasing} for the oracle case. The recommendation regime decreases the degree of local consumption, but not as much as the oracle benchmark. Due to the correlation of values, the oracle consumption path exploits this and leads to the consumption of more similar products than in the case when $\rho = 0$. However, since these spillovers also impact user learning in the no recommendation case, users \textit{over-exploit} this and increasingly consume products similar to good products that they have consumed before. This is further illustrated by Figure 3 in the supplementary material  which shows how the consumption paths in the oracle and no recommendation regimes vary as $\rho$ increases.
\par

Finally, this effect is further amplified as the level of risk aversion, $\gamma$, increases. Figure 4 in the supplementary material shows how drastically the degree of local consumption increases as $\gamma$ increases. This is due to the fact that the spillovers not only affect the mean expected belief about quality but also the degree of uncertainty. Local consumption therefore leads to users having less uncertainty about certain areas of the product space and risk aversion may lead them to increasingly consume nearby products.

@@

\begin{finding}\label{finding_diversity}
The impact of recommendation on product diversity:
\begin{enumerate}
\item When $\rho = 0$, product diversity is the same across all three recommendation regimes;
\item When $\rho > 0$, product diversity decreases across all recommendation regimes but decreases the most in the no recommendation regime. This effect is amplified as $\rho$ increases as well as when users become more risk-averse.
\end{enumerate}
\end{finding}
\par 
Finding~\ref{finding_diversity} summarizes the main results on product diversity. As before, when there is no correlation between valuations, product diversity is the same across different recommendation regimes. The over-exploitation of information spillovers when $\rho > 0$ leads to product diversity being lowest in the no recommendation regime. As a result, this effect gets amplified as $\rho$ increases, which leads to the gap in diversity between the regimes to increase as $\rho$ increases. Figure \ref{fig:diversity_welfare_correlation} shows how diversity varies as $\rho$ increases across the three regimes that we consider.There is a similar increasing diversity gap as $\gamma$, or the level of risk-aversion, increases as can be seen in Figure 4 in the supplementary material. The mechanisms behind these effects have direct parallels to those discussed in the previous section since low average sequential consumption distance directly leads to low diversity.
\par 
We now study how recommendation impacts user welfare in our model. In our model users make consumption decisions that maximize their current period ex-ante utility that depends on their beliefs in that period. Thus, from an ex-ante perspective, they make optimal decisions, but our primary interest is in understanding how the ex-post, or realized, utility varies across regimes and parameter values. We define user's \textit{ex-post} welfare as the average of the realized values, controlling for the effect of $T$:
$$W_i:= \frac{1}{T}\sum_{n \in C_i^T} x_{i,n}$$

\begin{finding}\label{finding_welfare_gap}
The impact of recommendation on consumer welfare is as follows:
\begin{enumerate}
\item Under oracle recommendation, welfare is constant as $\rho$ increases.
\item Under no recommendation, welfare is increasing as $\rho$ increases.
\item Recommendation introduces welfare gains relative to no recommendation, but these gains are decreasing as $\rho$ increases. 
\end{enumerate}
\end{finding}
\par 
Finding~\ref{finding_welfare_gap} states our main findings of the impact of recommendation on ex-post welfare, which are drawn from Figure 1 in the supplementary material. The most interesting observation is that the value of recommendation decreases as $\rho$ decreases. The intuition is clear as recommendation provides users with information that allows them to better guide their decisions and increase welfare. However, as $\rho$ increases users get increasingly more information from consuming items since the realized utility is now more informative about the utility of nearby items. Thus, since consumption decisions themselves reveal information, the information that recommendation reveals provides less value.
\par 
Figure 1 in the appendix shows how diversity and welfare vary across recommendation regimes as the the strength of the common-value component, $\beta$, varies. As expected, when $\beta = 0$, these values are the same. However, as $\beta$ grows the idiosyncratic component of user valuations becomes increasingly less important and the recommendation regime begins to converge to the oracle regime. Indeed, we see that both product diversity and welfare increase under recommendation, whereas under no recommendation we see that there is a decrease in diversity with little changes to consumer welfare. This is intuitive as $\beta$ controls the informativeness of recommendation about user preferences.
\par 
One striking observation is that the decrease in diversity does not appear to be associated with a decline in welfare. Indeed, it appears that the opposite is the case - that low diversity is associated with higher welfare and vice versa. Thus, we next explore the relationship between welfare and diversity.

\addtocounter{figure}{-1}
\begin{figure}[ht]
\caption{Diversity vs. Welfare, \textbf{No Recommendation}}
\begin{subfigure}{.45\textwidth}
\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn.pdf}
\end{subfigure}
\begin{subfigure}{.45\textwidth}
\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra.pdf}
\end{subfigure}\\
\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under the no recommendation regime. The figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.}\label{fig:diversity_welfare_ra}
\end{figure}

\addtocounter{figure}{-1}
\begin{figure}[ht]
\caption{Diversity vs. Welfare, \textbf{Recommendation}}
\begin{subfigure}{.45\textwidth}
\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_rn_partial.pdf}
\end{subfigure}
\begin{subfigure}{.45\textwidth}
\includegraphics[width=1.0\linewidth]{figures/diversity_welfare_ra_partial.pdf}
\end{subfigure}\\
\caption*{\scriptsize Notes: The figure plots the relationship between diversity and welfare under the recommendation regime. The figure on the left displays this relationship for when $\gamma = 0$ and the figure on the right for when $\gamma = 5$.}\label{fig:diversity_welfare_ra_partial}
\end{figure}
\addtocounter{figure}{-1}

@@

Finding~\ref{finding_diversity_welfare_corr} summarizes our findings on the relationship between diversity and welfare. Figure \ref{fig:diversity_welfare_ra} shows how diversity and welfare correlate for the no recommendation case as we vary the degree of risk aversion. When there is no risk-aversion then there is a negative correlation between welfare and diversity. This is since, with no risk-aversion, in a given period a user will select the good that she currently believes has the highest expected value. High product diversity in this case can arise from a user who consumes an item she disliked and updates her beliefs about nearby items negatively. As a result, in the following period she will pick an item far away in the product space from the item that was previously consumed. If instead the user valued highly the item that she had consumed, then she is more likely to pick a nearby item. The information spillovers therefore lead to high product diversity being negatively correlated with welfare.
\par
This only happens since $\gamma = 0$ leads to users only caring about the expected value of the product. However, as we saw in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, increasing $\gamma$ can lead to lower diversity and increasingly local consumption due to the fact that the degree of uncertainty now impacts users' choices. This weakens the negative relationship between diversity and welfare since both negative and positive experiences with a good reduce uncertainty about surrounding items. This leads to the inverted-U shape found in Figure \ref{fig:diversity_welfare_ra} when $\gamma$ is relatively large (e.g $\gamma = 5$) though diversity and welfare are virtually uncorrelated in the data. In the recommendation and oracle regimes, under risk neutrality ($\gamma=0$), welfare and diversity and uncorrelated, while under risk aversion ($\gamma=5$), it is possible to observe in Figure \ref{fig:diversity_welfare_ra_partial} an actual positive relation between diversity and welfare as recommendations are able to reduce uncertainty and facilitate exploration of the product space.

\subsection{User Homogenization}

\begin{figure}[ht]
\caption{Relationship between $\beta$ and Homogeneity, $N = 200$}
\includegraphics[width=.45\linewidth]{figures/beta_homogeneity_N_200_T_20}\label{fig:beta_homo}
\caption*{\scriptsize Notes: This figure displays the value of the homogeneity measure as we vary the weight of the common value component, $\beta$. Each line represents this plot for a single recommendation regime.}
\end{figure}

In this section, we focus on comparisons across users and investigate how the consumed set of items across users varies across different recommendation regimes and parameter values. In particular we look at the degree of \textit{homogenization} between users. Similar to other papers that study the degree of homogenization in recommender systems (e.g. \cite{chaney2018algorithmic}) we measure homogeneity via the Jaccard index between the consumption sets of users:
\begin{align*}
H:=\frac{1}{|I|(|I|-1)}\sum_{i,j \in I: i \ne j}d_J(C_i^T,C_j^T)
\end{align*}
where $d_J$ denotes the Jaccard index and $H \in [0,1]$.
\begin{finding}\label{finding_homogeneity}
The impact of recommendation on homogeneity is as follows:
\begin{enumerate}
\item Highest under recommendation and lowest under no recommendation;
\item Increasing in $\beta$, or the weight of the common-value component;
\item Decreasing in $\rho$ for partial recommendation, but weakly increasing in $\rho$ for no recommendation.
\end{enumerate}
\end{finding}
Finding~\ref{finding_homogeneity} summarizes our findings on the impact of recommendation on user homogeneity. First, we study how the degree of homogenization varies as we increase $\beta$, the weight of the common value component. As $\beta$ increases we expect that users should become increasingly homogeneous as the realized utilities of the items are now increasingly similar. Figure \ref{fig:beta_homo} confirms that as the weight of the common-value component, $\beta$, increases users consume increasingly similar sets of items. The homogenization effect is strongest under the recommendation regime since the revelation of the common-value component induces users to consume products in similar areas of the product space. As $\beta$ increases, some amount of homogenization is optimal as can be seen from the oracle case. However, since users in the no recommendation regime do not know the common-value component they engage in local consumption in different areas of the product space which leads to less than optimal homogeneity.
\par
We next study how the degree of homogeneity varies as $\rho$ increases. Figure 2 in the supplementary material shows how homogeneity varies as $\rho$ increases across the different recommendation regimes. The most striking finding is that homogeneity decreases as $\rho$ increases in the recommendation regime. As was highlighted in Findings \ref{finding_local_consumption} and \ref{finding_diversity}, the degree of local consumption increases with $\rho$. Even though the revelation of the common-value component induces them to search in similar parts of the product space, their idiosyncratic components induce them to consume products in a more localized area of the product space as $\rho$ increases which leads to a decline in homogeneity.

@@

In this section we discuss how the insights from our model of user decision-making can inform the evaluation and design of recommender systems. The classic approach to recommender system evaluation is to predict user ratings for items and to compare how accurate this prediction is to recorded ratings data, either explicitly given by users or inferred from behavioral data. The recommender system should then recommend the items with the highest predicted ratings \cite{adomavicius2005toward}.
